---
title: "p8105_hw2_anh2179"
author: Amitra Hoq
output: github_document
date: "2026-10-02"
---

# Problem 0
```{r, message=FALSE}
library(tidyverse)
library(readxl)
```


# Problem 1
## Importing and Cleaning NYC Transit Dataset
```{r}
transit_df = 
  read_csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) |>
  
  janitor::clean_names() |>
  select(
    line,station_name,station_latitude,station_longitude,starts_with("route"),
         entry,vending,entrance_type,ada) |>
  mutate(entry = recode_values(entry,"YES" ~ TRUE,"NO" ~ FALSE))
dim(transit_df)
```
### Paragraph about NYC Transit Dataset
This dataset describes entrances and exits at New York City subway stations. It includes each station's line, name, and location (latitude and longitude), the routes it serves (route1–route11), and details about each entrance: whether it allows entry, whether it has a vending machine, its entrance type, and whether it is ADA compliant. I imported the data, cleaned the variable names, and kept only the columns needed for later parts of this problem. I then converted entry from yes/no to a logical variable. I decided to also denote route8–route11 as characters so they match route1–route7. The resulting dataset has 1,868 rows and 19 columns.
The data is not tidy. Route number and route name should each be a single variable, but they are spread across 11 columns.

## Questions about NYC Transit Data  
```{r}
# How many distinct stations are there? 
transit_df |>
  select(station_name, line) |>
  distinct() |>
  nrow()

# How many stations are ADA compliant?
transit_df |>
  filter(ada == TRUE) |>
  select(station_name, line) |>
  distinct() |>
  nrow()

# What proportion of station entrances / exits without vending allow entrance?
transit_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```

## Reformatting NYC Transit Data
```{r}
transit_tidy = 
  transit_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route_name",)
```

## Questions about Reformatting NYC Transit Data 
```{r}
# How many distinct stations serve the A train?
transit_tidy |>
  filter(route_name == "A") |>
  select(station_name, line, route_name) |>
  distinct() |>
  nrow()

# Of the stations that serve the A train, how many are ADA compliant?
transit_tidy |>
  filter(route_name == "A",ada == TRUE) |> 
  select(station_name, line, ada, route_name) |>
  distinct() |>
  nrow()
```

# Problem 2
## Importing, Cleaning, Organizing and Combining Trash Wheel Datasets 
```{r}
mr_trash_df =
  read_excel(
    "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N743") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year= as.numeric(year), 
    plastic_bottles = as.double(plastic_bottles),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel")

professor_trash =
  read_excel(
    "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M140") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year= as.numeric(year),
    plastic_bottles = as.double(plastic_bottles),
    trash_wheel = "Professor Trash Wheel")

gwynnda =
  read_excel(
    "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395"
    ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year= as.numeric(year),
    plastic_bottles = as.double(plastic_bottles),
    trash_wheel = "Gwynnda"
    )

trash_wheel_df = bind_rows(mr_trash_df,professor_trash,gwynnda)
```

### Paragraph about Trash Wheel Data 

The combined `trash_wheel_df` dataset consists of `r nrow(trash_wheel_df)` observations and includes data from `r trash_wheel_df |> pull(year) |> min(na.rm = TRUE)` through `r trash_wheel_df |> pull(year) |> max(na.rm = TRUE)`. Some key variables include `year`, `sports_balls`, which was rounded and converted to an integer variable, and `plastic_bottles`, which is stored as a double variable. Given the data, the total weight of trash collected by Professor Trash Wheel is `r trash_wheel_df |> filter(trash_wheel == "Professor Trash Wheel") |> pull(weight_tons) |> sum(na.rm = TRUE)` tons. The total number of cigarette butts collected by Gwynnda in June of 2022 is `r trash_wheel_df |> filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |> pull(cigarette_butts) |> sum(na.rm = TRUE)`.

# Problem 3
## Import, Clean, and Tidy Baseline Demographics
```{r}
mci_baseline=
  read_csv("MCI_baseline.csv", show_col_types = FALSE, 
    skip = 1,
    na = c(".", ",","NA")) |>
  janitor::clean_names() |>
  mutate(sex = recode_values(sex,1 ~ "Male",0 ~ "Female"),
         apoe4 = recode_values(apoe4,1 ~ "Carrier",0 ~ "Non-carrier")) |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)
mci_baseline
dim(mci_baseline)
```
### Paragraph about Tidying Baseline Data 
To import the baseline data, we skip the first row of the CSV because it holds variable descriptions, not data. We treat `.` and `NA` as missing values, which lets `age_at_onset` be read as numeric. We clean the variable names with `janitor::clean_names()`. We then recode `sex` (1 = Male, 0 = Female) and `apoe4` (1 = Carrier, 0 = Non-carrier) from numeric codes to labeled categories, so they are not treated as numbers. To apply the inclusion criteria, we keep participants who never developed MCI (missing `age_at_onset`) and those whose onset came after their baseline age. Participants whose onset was at or before baseline already had MCI when they entered the study, so they are removed. The cleaned dataset has `r nrow(mci_baseline)` rows and `r ncol(mci_baseline)` columns: `r paste(names(mci_baseline), collapse = ", ")`. In total, 483 participants were recruited, and `r nrow(mci_baseline)` remain after applying the inclusion criteria. Of these, `r mci_baseline |> filter(!is.na(age_at_onset)) |> nrow()` developed MCI during follow-up. The average baseline age is `r mci_baseline |> pull(current_age) |> mean(na.rm = TRUE)` years. Among women in the study, the proportion who are APOE4 carriers is `r mci_baseline |> filter(sex == "Female") |> summarize(prop = mean(apoe4 == "Carrier")) |> pull(prop)`.




## Import, Clean, and Tidy Amyloid Biomarker Data
```{r}
amyloid_df =
  read_csv("mci_amyloid.csv",
    skip = 1,
    na = c("Na", ".", ",")
  ) |>
  janitor::clean_names() |>
  rename(
    id = study_id,
    baseline_years = baseline
  )

amyloid_df
dim(amyloid_df)
```

### comment on the steps on the import process and the features of the dataset.

## Checking 
```{r}
mci_baseline |>
  anti_join(amyloid_df, by = "id") |>
  nrow()
 amyloid_df |>
  anti_join(mci_baseline, by = "id") |>
  nrow()
```
### Paragraph on findings 

## Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained
```{r}
mci_df =
  inner_join( mci_baseline,amyloid_df,by = "id")
```
### Paragraph 

## Exporting CSV to directory 
```{r}
write_csv(mci_df,"mci_data.csv")
```


